Early Diagnosis and Prognostic Prediction of Colorectal Cancer through Plasma Methylation Regions

Version 1

Abstract

Abstract

The methylation of plasma cell-free DNA (cfDNA) has emerged as a valuable diagnostic and prognostic biomarker in various cancers including colorectal cancer (CRC). Currently, there are no biomarkers that serve simultaneously for early diagnosis and prognostic prediction in CRC patients. Herein, we developed a plasma panel (27 DMRs, differential methylated regions) and validated its superior performance across CRC diagnosis and prognosis prediction in an independent cohort. We first conducted a preliminary screening of 119 CRC tissue samples to identify CRC-specific methylation features. Subsequently, a CRC-specific methylation panel was developed by further filtering 161 plasma samples. Then machine learning algorithms were applied to develop diagnosis and prognosis models using cfDNA samples from 51 CRC patients and 33 normal controls. The diagnosis model was tested in a cohort consisting of 30 CRC, 37 advanced adenoma (AA), and 14 healthy plasma samples, independently validated in a cohort consisting of 18 CRC, 91 NAA, 23 AA and 34 healthy plasma samples. In the tissue external validation cohort (GSE48684), the cfDNA methylation diagnosis model conducted with the panel, have the area under the curve (AUC) reached 0.983, and for the plasma cfDNA model in the external validation cohort, the sensitivities for NAA, AA and CRC 0 -Ⅱ are 48.4%. 52.2% and 66.7% respectively, with a specificity of 88%. Additionally, the panel was applied to patient staging and metastasis, performing well in predicting CRC distant metastasis (AUC = 0.955) and prognosis (AUC = 0.867). Using normal samples as control, the changes in methylation score in both tissue and plasma were consistent across different lesions, although the degree of alterations varied with severity. The methylation scores vary between paired tissue and blood samples, suggesting distinct mechanisms of migration from tumor tissue to blood for the 27 DMRs. Together, Our cfDNA methylation models based on 27 DMRs can identify different stages of CRC and predict metastasis and prognosis, ultimately enabling early intervention and risk stratification for CRC patients.

Introduction

Colorectal cancer (CRC) is the third most prevalent cancer globally and ranks as the fourth leading cause of cancer death, accounting for 9.4% of cancer-related mortalities (1). When patients were diagnosed with CRC, over half of their family (62.9%) faced financial burdens (2, 3). Extensive research underscores that the survival rates of individuals diagnosed with advanced CRC (Stage III or IV) witness a significant decrement (4, 5). Early detection and removal of precancerous lesions remain the most effective strategy to prevent CRC-associated mortality (6). While colonoscopy is regarded as the gold standard for CRC detection, its limitations including invasiveness, suboptimal patient compliance and risk of complications including intestinal perforation, warrant further consideration (7). Guaiac-based fecal occult blood test (gFOBT), fecal immunochemical test (FIT), carcino-embryonic antigen (CEA), and multi-target stool DNA (mt-sDNA) testing, are constrained in clinical application due to the lower sensitivity(8, 9).

Recent research demonstrated a profound correlation between CRC and the development of genetic and epigenetic alterations. During early stage of CRC development, epigenetic modifications surpass the frequency of gene mutations, indicating their potential as diagnostic biomarkers in screening for colon polyps and cancer(10). Circulating tumor DNA (ctDNA) in cell-free DNA (cfDNA) primarily come from apoptotic and necrotic tumor cells, carrying cancer-specific epigenetic alterations (11). Notably, blood cfDNA methylation emerges as a promising cancer screening pathway because of its early appearance in tumorigenesis and abundant signal density (12).

The value of cfDNA methylation in early diagnosis, detection of recurrence, molecular subtyping and prognostic prediction of CRC has been proven(13). However, there are still reported limitations. One study found a diagnostic panel for CRC demonstrated great performance, but it was derived from CRC tissue and normal blood leukocyte methylation data, potentially introducing bias due to the inconsistent sample types (14). Another study addressed the potential of cfDNA methylation in patient risk stratification, but the requirement for blood sampling every three months led to poor patient compliance(15).

In this study, we aimed to develop plasma biomarkers derived from CRC tissues that exhibit superior performance in the diagnosis, metastasis and prognostic prediction for CRC. We meticulously screened CRC and normal tissues, and refining our selection within plasma samples. More importantly, the substantial proportion of paired tissue and plasma samples (from the same patient), effectively minimized potential data bias. The final resulting panel comprised 27 differential methylated regions (DMRs), with the area under the curve (AUC) reached 0.983 in the tissue validation cohort. Given that DNA methylation markers in plasma primarily stem from tumor tissues, these 27 DMRs exhibit significant potential in CRC plasma diagnostics. With a single blood test, we can ascertain the presence of CRC, distinguish the specific stages of the lesion (advanced adenoma, CRC 0-II, CRC III-IV), help identify distant metastasis, and achieve optimal risk stratification for CRC patients. This test is of clinical significance and holds promise in guiding the diagnosis and treatment for CRC.

Materials and Methods

Study design and samples

Specific DNA methylation markers for CRC were identified through analysis of TCGA public database data, alongside collected tissue and plasma samples. Subsequently, diagnosis, metastasis, and prognosis models for CRC were established and validated. The 450k chip methylation data encompassing colorectal, esophageal, gastric, lung, liver, and breast cancers, and CRC transcriptome sequencing data were downloaded from The Cancer Genome Atlas (TCGA) database. The methylation data of tissue validation cohort was obtained from

Fifteen pairs of tissue samples from the Seventh Medical Center of PLA General Hospital were sequenced with the Roche Nimble Gen Seq Cap Epi method. The 89 tissue samples and 77 plasma samples used for marker screening were collected between June and December 2020 at the Seventh Medical Center of PLA General Hospital and Dongying People’s Hospital. Participants for model establishment and validation were enrolled from June 2020 to July 2022 in the Seventh Medical Center of PLA General Hospital, Dongying People’s Hospital.

The paeticipants for external validation cohort were enrolled from June 2021 to December 2022 at the First Hospital of Longyan, Fujian Medical University. Inclusion and exclusion criteria are outlined in the Supplementary Materials. Detailed methods for DNA extraction, library construction, targeted bisulfite sequencing, and methylation data processing are provided in the Supplementary Materials.

CRC-specific methylation markers selection

Combining TCGA methylation data with our 15 pairs of tissue methylation sequencing results, probes were designed for further screening in the 89 samples (13 normal tissues and 38 paired CRC tissues), 404 DMRs were identified. The detailed methods were listed in supplementary methods. Then, the methylation levels of 77 plasma samples from 13 Normal, 15 non-advanced adenoma (NAA), 12 advanced adenoma (AA) and 37 CRC participants, were analyzed. Subsequently, the Least Absolute Shrinkage and Selection Operator (LASSO) regression analysis was used to filter markers, and 27 regions that appeared in more than 50 iterations were chosen as plasma diagnostic markers. The relationship between the methylation levels of the 27 DMRs and the transcription levels of their respective genes was compared by Spearman correlation coefficients.

Construction of the CRC diagnosis model

The methylation data from the training cohort underwent machine learning, culminating in the establishment of a five-fold cross-validated model characterized by binary deviance minimization standards. The diagnosis model was constructed using logistic regression, and its performance was evaluated through Receiver Operating Characteristic (ROC) curve analysis. The optimal cutoff value was determined using the maximum Youden index. The methylation data of tissue validation cohort (GSE48684) was downloaded from GEO (Gene Expression Omnibus) dataset. Using the methylation scores derived from the diagnosis model, the CRC staging (AA, CRC 0-Ⅱ, CRC Ⅲ-Ⅳ) was predicted.

Construction of the CRC metastasis and prognosis models

Utilizing the methylation scores derived from the diagnosis model, in conjunction with patient pathological parameters (metastasis), the plasma CRC metastasis model was constructed. The performance of this model underwent evaluation through ROC analysis, with the optimal cutoff value determined via the maximum Youden index. Furthermore, employing the same methodology, the plasma prognosis model was established and validated, based on the methylation scores derived from the CRC diagnosis model and survival information from the patients.

Statistical analysis

All statistical analyses were conducted using SPSS 26.0 or R 4.1.3. Two-sided tests were used for p-values, with differences deemed statistically significant at P < 0.05. Model performance was assessed through ROC analysis using the “roc” function from the R package pROC, generating AUC and a 95% confidence interval (CI). Kaplan-Meier (KM) curves were used for survival curve estimation, with comparisons made through the log-rank test and hazard ratios determined by Cox regression.

Results

Patient characteristics

To delineate DNA methylation biomarkers specific to CRC, 119 tissues and 77 plasma samples (originating from the same patients with tissues) were collected for methylation sequencing analysis, narrowing down to 142 DMRs. Then the 27 DMRs were selected with the sequencing data of 84 plasma samples. Following this, CRC diagnosis, metastasis, and prognosis models were constructed using plasma samples from 84 individuals (51 CRC, 33 Normal). Validation was then carried out in an independent cohort (30 CRC, 37 AA, 14 Normal). The CRC diagnosis model was further tested in an external validation cohort (18 CRC 0-II, 91 NAA, 23 AA, 34 Normal). Detailed information about the study design is illustrated in Figure 1, and comprehensive patient characteristics are summarized in Table s2-4.

Flow diagram of the study.

Methylation markers selection

Analyzing the 450k chip methylation data from TCGA, the differential methylated sites (DMCs) in CRC were identified (Figure s1A). Combining with additional CRC-related methylation sites reported in the literature, 1438 DMCs were selected for validation. The heat map depicted the methylation levels of 1438 DMCs, revealing distinct methylation patterns between Normal and Tumor. (Figure s1B). The top 1400 DMCs were carefully selected based on the results of DNA methylation sequencing of 15 paired tissue samples (Figure s2A-C).

The distribution of the DMCs was predominantly observed in introns and promoters, signifying substantial implications for gene expression and cellular functions (Figure s2D).

Merging the 1438 DMCs with the 1400 DMCs for probe design, 404 DMRs were discerned across 89 tissues (Figure s3A). The different methylation patterns between Normal and Tumor were also quite evident (Figure s3B-C). These DMRs correspond to genes that play pivotal roles in biological processes such as cell-cell adhesion, digestive system development, and cAMP and cGMP pathways (Figure s3D). Subsequently, 142 regions remained after sequencing 77 plasma samples (Figure s3E-F). The final 27 DMRs were selected after analyzing methylation data of the training cohort samples. Detailed information of the 27 DMRs is provided in Table s5. Correlation analysis revealed the methylation levels of the majority of the 27 DMRs are correlated with the transcription levels of their respective genes (Figure s4).

Development and validation of the CRC diagnosis model

Using methylation data from 27 DMRs in tissues, we applied machine learning to construct a diagnosis model for CRC and the AUC reached a high value of 0.994 (Figure 2A). In the external tissue independent validation cohort (GSE48684), the AUC for CRC is 0.983 and for AA is 0.966. (Figure 2B-C). The sensitivities for AA and CRC were 95% and 94%, respectively (Figure 2D-E). Besides, the plasma-based diagnosis model for CRC was constructed with training cohort and validated in an external independent validation cohort, the 27 DMRs exhibited distinctive methylation patterns between CRC and normal individuals (Figure 3A-B). The methylation scores of the model were significantly elevated in the CRC group compared to the normal and AA groups (P< 0.001), and they increased with advance tumor staging (Figure 3H). In the training cohort, the ROC curve analysis showed an AUC value of 0.928 [95% confidence interval (CI): 0.874–0.981] (Figure 3C). The sensitivities of the CRC diagnosis model in the training cohort for stages 0-II, III-IV, and all CRC stages were 80%, 92%, and 86%, respectively, with a specificity of 85% (Figure 3I). ROC curve analysis in the validation cohort showed AUC values for AA, stages 0-II, III-IV, and all CRC stages as 0.714 (0.550–0.878), 0.890 (0.749–1.000), 0.967 (0.899–1.000), and 0.929 (0.827–1.000). (Figure 3D-G). The CRC diagnosis model demonstrated a specificity of 93% in the validation cohort, with sensitivities for AA, stages 0-II, III-IV, and all CRC stages of 43%, 67%, 100%, and 83%, respectively (Figure 3I). We further tested the early diagnostic performance of the model in an external validation cohort, and the results showed that the model achieved a sensitivity of 52% for AA and 48% for NAA (Figure 4A-E), The plasma diagnosis model also serves to discern the specific staging of CRC patients. ROC curve analysis on the validation cohort revealed an AUC of 0.813 (0.656–0.970) for discriminating between CRC 0-II and CRC III-IV, and an AUC of 0.799 (0.693–0.905) for distinguishing between AA and CRC (Figure s5A-C). Kaplan-Meier survival curves revealed a significant reduction in overall survival (OS) for CRC III-IV patients compared to CRC 0-II patients identified by this diagnosis model (Figure s5D-E).

Development and validation of tissue diagnosis model based on the 27 CRC-specific DMRs.

(A) The use of ROC curve analysis to assess the performance of the tissue diagnostic model in differentiating CRC patients from normal individuals of the training cohort (51 tissues we collected). (B-C) The use of ROC curve analysis to assess the performance of the tissue diagnostic model in differentiating CRC patients from normal individuals(B) and distinguish AA from Normal subjects (C) in tissue validation cohort 1(GSE 48684 dataset). (D-E) The sensitivity and specificity of the diagnosis model in the tissue training(D) and validation cohort 1(E).

Development and validation of a plasma diagnosis model based on the 27 CRC-specific DMRs.

(A-B) Heatmap illustrating the DMRs between CRC and advanced adenoma and healthy controls in the training (A) and validation cohort (B). (C) The use of ROC curve analysis to assess the performance of the diagnosis model in differentiating CRC patients from normal individuals of the training cohort. (D) The use of ROC curve analysis to evaluate the performance of the diagnosis model in differentiating CRC patients from normal individuals of the validation cohort. (E-G) In the validation cohort, the performance of the diagnosis model in differentiating early-stage CRC (0-Ⅱ) from normal individuals (E); in differentiating advanced-stage CRC (Ⅲ-Ⅳ) from normal individuals (F); in differentiating AA from normal individuals (G). (H) Methylation scores of Normal AA and CRC individuals generated from the diagnosis model in the training and validation cohort (The dashed line represents the cutoff value). (I) The sensitivity and specificity of the diagnosis model stratified by stages in the training and validation cohort. Abbreviations: DMRs, differential methylated regions; ROC, receiver operating characteristic; AA, advanced adenoma; CRC, colorectal cancer.

External validation of plasma diagnosis model.

(A-C) In the external validation cohort, the performance of the diagnosis model in differentiating early-stage CRC (0-II) from normal individuals (A); in differentiating AA from normal individuals (B); in differentiating NAA from normal individuals (C). (D) Methylation scores of Normal, NAA, AA and CRC (0-II) individuals generated from the diagnosis model in the external validation cohort (The dashed line represents the cutoff value). (E) The sensitivity and specificity of the diagnostic model for NAA, AA, and CRC in the external validation cohort.

Development and validation of the CRC metastasis and prognosis models

Through an analysis of the relationship between the methylation scores in the diagnosis model of CRC patients’ plasma samples and clinical-pathological parameters, we observed that methylation scores are associated with metastasis and staging (P< 0.001, Figure 5A-B), but not with age, gender, and lesion location (P>0.05, Figure s6A-C). This suggests the possible role of 27 DMRs in metastasis and prognosis prediction. Therefore, we developed plasma-based CRC metastasis and prognosis models based on the 27 CRC-specific DMRs. In the training and validation cohorts, the metastasis model demonstrated high AUCs of 0.969 (95% CI: 0.926-1) and 0.955 (95% CI: 0.878-1), respectively (Figure 5C). Methylation scores of M1-CRC individuals identified from the metastasis model were significantly higher than M0-CRC individuals (Figure s6D). Additionally, the M1-CRC patients exhibited shorter OS than M0-CRC patients. (Figure s6E-F). For the prognosis model, ROC curves demonstrated excellent performance in the training cohort (AUC = 0.883, 95% CI: 0.778-0.988) and validation cohort (AUC = 0.867, 95% CI = 0.728-1). (Figure 5D). Furthermore, Kaplan-Meier survival curves revealed a significant reduction in OS for the high-risk group of CRC patients identified by this prognosis model (Figure 5E). Based on the cutoff value determined by the algorithm, we divided CRC patients into high-risk and low-risk groups. Notably, almost all patients who developed distant metastases and those who had died belonged to the high-risk group (Figure 5F, H). We further characterized the survival status distribution and metastasis status between the two groups. As expected, the high-risk group had a higher proportion of deceased individuals, while non-metastatic patients were more prominent in the low-risk group (Figure 5G, I). These results indicate that the metastasis and prognosis models successfully identified patients who require further treatment. Multivariate regression analysis revealed a substantial correlation between the methylation score of the 27 DMRs and OS, indicating methylation score as an independent prognostic factor for CRC (Table s6). These findings underscore the considerable potential of the prognosis model based on the 27 DMRs specific to CRC in predicting the prognosis and conducting risk stratification for CRC patients.

Development and validation of a plasma-based metastasis model and a plasma-based prognosis model.

(A-B) The methylation scores generated by the diagnosis model for CRC patients across various M stages (A) and different disease stages (B). (C) ROC curve analysis was performed to assess the performance of 27 DMRs methylation scores generated by the diagnosis model in distinguishing M1-CRC patients from M0-CRC patients in the training (AUC = 0.969) and validation (AUC = 0.955) cohorts. (D) ROC analysis evaluates the performance of 27 DMRs methylation scores generated by the diagnosis model in predicting the prognosis of CRC patients. (E) Kaplan-Meier survival curves comparing OS between the high-risk CRC group and low-risk CRC group of prognosis model in the training cohort and validation cohort. (F, H) Metastasis and prognosis prediction of the training and validation patitients(n = 88). The black dotted line at the cutoff value divides the patients into high-risk and low-risk groups. Yellow circles and gray circles represent patients without distant metastasis and with distant metastasis, respectively (F). Yellow circles and gray circles represent patients survived and deceased, respectively (H). (G, I) The proportion of metastasis and deceased is higher in the high-risk group. The p-value was calculated using a two-sided Fisher’s exact test. Abbreviations: DMRs, differential methylated regions; ROC, receiver operating characteristic; M1-CRC, colorectal cancer with distant metastasis; M0-CRC, colorectal cancer without distant metastasis; AUC, the area under the curve.

Changes in methylation scores of tissue and plasma samples across different populations

Collectively, the CRC diagnosis model, established based on the methylation levels in these 27 DMRs within tissues, exhibits robust performance. However, the plasma-based diagnosis model exhibited reduced performance in AA and CRC 0-II compared to CRC III-IV. Therefore, we analyzed the methylation scores of the 27 DMRs generated from diagnosis models in paired tissue and plasma samples. Notably, in tissues, the methylation scores are ranked as AA > CRC > Normal (Figure 6A), suggesting significant alterations in the methylation of the 27 DMRs at the onset of precancerous lesions. In plasma, methylation scores gradually increase with CRC progression (CRC III-IV > CRC 0-II > AA > Normal, Figure 6B). Further analysis of individual DMR methylation levels in the blood revealed discernible distinctions in certain DMRs during both AA and CRC 0-II stages, while variances in another subset of DMRs became detectable exclusively during CRC III-IV stages (Figure 6C). We then conducted a gene analysis of DMRs identifiable in early-stage CRC blood on the Metascope website. Our findings revealed their association with processes such as “secretion by cell” and “cell-cell adhesion.” (Figure s6G) Consequently, we hypothesize that although alterations of 26 DMRs were observed in AA tissues, only a selected few are actively released into the bloodstream through cell secretion by tumor cells. The majority of these alterations are likely to be identified in the blood only after the occurrence of apoptosis or necrosis in tumor cells. These results elucidate the inconsistency in the performance of tissue and plasma diagnosis models and the differences in diagnostic efficacy for different CRC stages using the same model.

Methylation scores of paired tissue and plasma samples.

(A) The methylation scores of tissues from Normal, AA, CRC 0-Ⅱ, CRC Ⅲ-Ⅳ individuals. (B) The methylation scores of blood samples from Normal, AA, CRC 0-Ⅱ, CRC Ⅲ-Ⅳ individuals. (C) The mechanism diagram interpreting the methylation scores changes of tissue and plasma across different populations.

Discussion

Liquid biopsy of cfDNA has become an ideal clinical detection method because it is minimally invasive and easily sampled. However, due to the low concentration of ctDNA in bodily fluids and the heterogeneity of tumor cells, many DNA markers for CRC currently lack certainty whether they originate from CRC tissues. In this study, we integrated TCGA tissue data with self-collected tissue methylation data to identify CRC-related methylation sites.

Probes were designed and applied in paired tissue and plasma samples to further screen the methylation sites. We successfully determined 27 DMRs originating from CRC tissues, which were subsequently used to construct diagnostic model. In the validation cohorts of tissue and plasma samples, the diagnosis models based on the 27 DMRs we established could effectively differentiate samples between CRC and normal participants, in addition, the plasma diagnosis model could distinguish the different stages of CRC. Furthermore, the transfer prognosis model, established based on diagnosis model scores, could effectively demonstrate whether the CRC had metastasized, separating CRC patients into high-risk and low-risk groups.

Several blood-based methylation biomarker candidates have been proposed for early detection of CRC. For instance, the FDA-approved circulating methylated SEPT9 DNA (mSEPT9) demonstrates sensitivities of 11.2%, 35.0%, 63.0%, 46.0%, and 77.4% for AA and CRC stages I-IV, respectively, with a specificity of 91.5% (16). In a recent study, a cfDNA methylation-based CRC screening model has a sensitivity of 86.4% and a specificity of 90.7%, utilizing 149 markers derived from blood samples (17). In our study, 27 markers were selected through a layered screening process from tissue and plasma, undergoing marker selection, model development, and validation to ensure model robustness. Then the superior performance of our plasma diagnosis model was validated, with a sensitivity of 83.3% and specificity of 92.9%. Notably, the sensitivity for AA and NAA reached 52.2% and 48.4% respectively, far surpassing the 11.2% sensitivity of mSEPT9 and the 33.3% reported in another blood screening study.(16, 17). In summary, our diagnostic model performs well in detecting precancerous lesions of CRC and has the potential to become a screening method for high-risk populations of CRC.

In plasma, we observed the methylation scores of 27 DMRs gradually increase with CRC progression. It is possibly due to less vascular infiltration in early CRC (18, 19). On the other hand, the ctDNA detected in the blood of early-stage CRC is largely derived from tumor cells actively secreting into the bloodstream, making it challenging to be precisely captured with current detection technologies due to the limited quantity. Conversely, ctDNA in the blood of late-stage CRC primarily emanated from apoptotic and necrotic tumor cells, resulting in a larger quantity that facilitates easier detection. However, in reality, the methylation scores of the 27 DMRs have indeed exhibited noticeable changes in AA and early cancerous tissues.

Additionally, in the tissue diagnosis model using the 27 DMRs, the sensitivity was 94% and the specificity was 100%. Therefore, we have grounds to believe that the 27 DMRs detected in plasma originate from CRC tumor tissues. With the implementation of more sensitive detection methods, the performance of these 27 DMRs in early CRC plasma diagnosis is likely to be further enhanced.

Additionally, cfDNA contributes to risk stratification and early recurrence detection in CRC(20). However, current methods rely on continuous blood cfDNA testing by patients (15, 21).

This study discovered that the preoperative cfDNA methylation level of 27 DMRs is linked to distant metastasis and is valuable for predicting the prognosis of CRC, consistent with certain prior research. (22, 23). Our model predicted AUCs of 0.955 and 0.867 for distant metastasis and prognosis in CRC, respectively. This underscores the potential utility of the preoperative application of this cfDNA methylation model for risk stratification, serving as an effective tool to improve the perioperative management of CRC patients.

The treatment strategies for CRC differ significantly across various stages. Molecular stratification approaches for CRC patients are on the rise, and concurrently, the clinical application of biomarkers to determine treatment decisions is gradually gaining traction (24). Studies have demonstrated the precise identification of T1 CRC patients at risk of lymph node metastasis using a specific set of miRNAs, thereby potentially mitigating unnecessary overtreatment (25, 26). Our CRC diagnosis model, with an AUC of 0.813 for distinguishing CRC 0-Ⅱ from CRC Ⅲ-Ⅳ, could also serve as a potent, straightforward, and cost-effective preoperative screening/detection method, guiding patients to select more appropriate treatment plans.

DNA methylation regulates gene transcription, guiding the progression from normal mucosa to AA and ultimately CRC. This process involves silencing tumor suppressor genes and activating oncogene transcription (27). Gene transcription levels correlate notably with methylation levels at specific sites, consistent with our research findings. Additionally, several genes housing the identified 27 DMRs have been partially explored in CRC. Some genes contribute to tumor growth. For instance, high methylation and diminished expression of transmembrane protein 240 (TMEM240) regulate CRC cell proliferation, predicting poor prognosis (28). The transcription factor homeobox A3 (HOXA3) activates aerobic glycolysis, promoting tumor growth (29). KIFC3 controls mitotic spindle assembly initiation (30). Moreover, several genes are implicated in tumor metastasis, such as NOVA alternative splicing regulator 1 (NOVA1), which promotes CRC migration by activating the Notch pathway (31). Protein tyrosine phosphatase receptor type T (PTPRT) contributes to early CRC dissemination (32).

SPARC-related modular calcium binding 2 (SMOC2) serves as the distinctive signature of cancer stem cells (CSCs) in CRC and promotes epithelial-to-mesenchymal transition (EMT) (33, 34). Increased methylation and expression of pancreatic and duodenal homeobox 1 (PDX1) facilitate CRC invasion and migration(35). However, the mechanisms by which alterations in specific DNA methylation impact gene expression are intricate. Certain transcription factors selectively recognize sequences with methylated CpG (mCpG) and influence the expression of multiple genes(36). Further exploration of the potential functional mechanisms of these DNA methylation markers may deepen our understanding of the molecular processes underlying CRC development, offering promising therapeutic targets.

This study has certain limitations. Firstly, although obtained from multiple institutions, the small sample size was insufficient. Therefore, the cfDNA methylation model should be further validated in large-sample trials in the future. Secondly, a prospective study is necessary to compare or combine the cfDNA methylation model with clinically commonly used markers such as CEA and CA19-9.

In conclusion, our cfDNA methylation model based on 27 DMRs can identify different stages of CRC, predict metastasis and prognosis, and ultimately achieve early intervention and risk stratification for CRC patients. The preoperative application of our DNA methylation biomarker as a robust, convenient, and cost-effective detection method can contribute to making more informed clinical decisions and improving the perioperative management of CRC patients.

eLife Assessment

This study presents an important finding that has identified 27 differentially methylated regions as a signature for non-invasive early cancer detection and predicting prognosis for colorectal cancer. The findings demonstrate promising clinical potential, particularly for improving cancer screening and patient monitoring. However, the evidence supporting the claims of the authors is incomplete due to a small sample size and some methodological concerns. The work will be of interest to researchers interested in cancer diagnosis or colorectal cancer monitoring.

Reviewer #1 (Public review):

Summary:

Colorectal cancer (CRC) is the third most common cancer globally and the second leading cause of cancer-related deaths. Colonoscopy and fecal immunohistochemical testing are among the early diagnostic tools that have significantly enhanced patient survival rates in CRC. Methylation dysregulation has been identified in the earliest stages of CRC, offering a promising avenue for screening, prediction, and diagnosis. The manuscript entitled "Early Diagnosis and Prognostic Prediction of Colorectal Cancer through Plasma Methylation Regions" by Zhu et al. presents that a panel of genes with methylation pattern derived from cfDNA (27 DMRs), serving as a noninvasive detection method for CRC early diagnosis and prognosis.

Strengths:

The authors provided evidence that the 27 DMRs pattern worked well in predicting CRC distant metastasis, and the methylation score remarkably increased in stage III-IV.

Weaknesses:

The major concerns are the design of DMR screening, the relatively low sensitivity of this DMR pattern in detecting early-stage CRC, the limited size of the cohorts, and the lack of comparison with the traditional diagnosis test.

Reviewer #2 (Public review):

This work presents a 27-region DMR model for early diagnosis and prognostic prediction of colorectal cancer using plasma methylation markers. While this non-invasive diagnostic and prognostic tool could interest a broad readership, several critical issues require attention.

Major Concerns:

(1) Inconsistencies and clarity issues in data presentation

a) Sample size discrepancies

- The abstract mentions screening 119 CRC tissue samples, while Figure 1 shows 136 tissues. Please clarify if this represents 119 CRC and 17 normal samples.

- The plasma sample numbers vary across sections: the abstract cites 161 samples, Figure 1 shows 116 samples, and the Supplementary Methods mentions 77 samples (13 Normal, 15 NAA, 12 AA, 37 CRC).

b) Methodological inconsistencies

- The Supplementary Material reports 477 hypermethylated sites from TCGA data analysis (Δβ>0.20, FDR<0.05), but Figure 1 indicates 499 sites.

- The manuscript states that analyzing TCGA data across six cancer types identified 499 CRC-specific methylation sites, yet Figure 1 shows 477. Please also explain the rationale for selecting these specific cancer types from TCGA.

- "404 CRC-specific DMRs" mentioned in the main text while "404 MCBs" in Figure 1, the authors need to clarify if these terms are interchangeable or how MCBs are defined.

(2) Methodological documentation

- The Results section requires a more detailed description of marker identification procedures and justification of methodological choices.

- Figure 3 panels need reordering for sequential citation.

(3) Quality control and data transparency

- No quality control metrics are presented for the in-house sequencing data (e.g., sequencing quality, alignment rate, BS conversion rate, coverage, PCA plots for each cohort).

- The analysis code should be publicly available through GitHub or Zenodo.

- At a minimum, processed data should be made publicly accessible to ensure reproducibility.

Reviewer #3 (Public review):

Summary:

This article provides a model for early diagnosis and prognostic prediction of Colorectal Cancer and demonstrates its accuracy and usability. However, there are still some minor issues that need to be revised and paid attention to.

Strengths:

A large amount of external datasets were used for verification, thus demonstrating robustness and accuracy. Meanwhile, various influencing factors of multiple samples were taken into account, providing usability.

Weaknesses:

There are notable language issues that hinder readability, as well as a lack of some key conclusions provided.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

Colorectal cancer (CRC) is the third most common cancer globally and the second leading cause of cancer-related deaths. Colonoscopy and fecal immunohistochemical testing are among the early diagnostic tools that have significantly enhanced patient survival rates in CRC. Methylation dysregulation has been identified in the earliest stages of CRC, offering a promising avenue for screening, prediction, and diagnosis. The manuscript entitled "Early Diagnosis and Prognostic Prediction of Colorectal Cancer through Plasma Methylation Regions" by Zhu et al. presents that a panel of genes with methylation pattern derived from cfDNA (27 DMRs), serving as a noninvasive detection method for CRC early diagnosis and prognosis.

Strengths:

The authors provided evidence that the 27 DMRs pattern worked well in predicting CRC distant metastasis, and the methylation score remarkably increased in stage III-IV.

Weaknesses:

The major concerns are the design of DMR screening, the relatively low sensitivity of this DMR pattern in detecting early-stage CRC, the limited size of the cohorts, and the lack of comparison with the traditional diagnosis test.

We sincerely thank the reviewer for their thorough evaluation and constructive feedback on our manuscript. We are encouraged that the reviewer found our 27-DMR panel promising for predicting distant metastasis and for its performance in late-stage CRC. We have carefully considered the weaknesses pointed out and have made revisions to address these concerns, which we believe have significantly strengthened our paper.

We agree with the reviewer that achieving high sensitivity for early-stage disease is the ultimate goal for any noninvasive screening test. Detecting the minute quantities of cfDNA shed from early-stage tumors is a well-recognized challenge in the field. Although the sensitivity of our current panel for early-stage CRC is modest, its core strengths, lie in its capability to also detect advanced adenomas and its excellent performance in assessing CRC metastasis and prognosis. Furthermore, we have now added a direct comparative analysis of our 27-DMR panel against the most widely used clinical serum biomarker for CRC, carcinoembryonic antigen (CEA), using samples from the same patient cohorts. Our results demonstrate that 27-DMR methylation score significantly outperforms CEA in diagnostic accuracy for early-stage CRC (64% vs. 18%) (Table s7). And in the Discussion section, we have also acknowledged our limitations and suggest that future studies are warranted to combine the cfDNA methylation model with commonly used clinical markers, such as CEA and CA19-9, with the aim of improving the sensitivity for early diagnosis.

We acknowledge the reviewer's concern regarding the cohort size and validation in larger, prospective, multi-center cohorts is essential before this panel can be considered for clinical application. We have explicitly stated this as a limitation of our study in the Discussion section and have highlighted the need for future large-scale validation studies (Page 18, Lines 367-373). We once again thank the reviewer for their insightful comments, which have allowed us to substantially improve our manuscript. We hope that the revised version is now suitable for publication.

Reviewer #2 (Public review):

This work presents a 27-region DMR model for early diagnosis and prognostic prediction of colorectal cancer using plasma methylation markers. While this non-invasive diagnostic and prognostic tool could interest a broad readership, several critical issues require attention.

Major Concerns:

(1) Inconsistencies and clarity issues in data presentation

(a) Sample size discrepancies

The abstract mentions screening 119 CRC tissue samples, while Figure 1 shows 136 tissues. Please clarify if this represents 119 CRC and 17 normal samples.

We sincerely thank the reviewer for this careful observation and for pointing out the inconsistency. We apologize for the error and the confusion it caused. Regarding Figure 1: The reviewer is correct. The number 136 in the original Figure 1 was an error. This was due to an inadvertent double-counting of the tumor samples that were used in the differential analysis against adjacent normal tissues. The actual number of tissue samples used in this analysis is 89. We have now corrected this value in the revised Figure 1.

Regarding the Abstract: The 119 CRC tissue samples mentioned in the abstract represents the total number of unique tumor samples analyzed across all stages of our study. This number is composed of two cohorts: the initial 15 pairs of tissues used for preliminary screening, and the subsequent 89 tissue samples used for validation, totaling 119 samples. We have ensured all sample numbers are now consistent throughout the revised manuscript.

The plasma sample numbers vary across sections: the abstract cites 161 samples, Figure 1 shows 116 samples, and the Supplementary Methods mentions 77 samples (13 Normal, 15 NAA, 12 AA, 37 CRC).

We sincerely thank the reviewer for their meticulous review and for identifying these inconsistencies in the plasma sample numbers. We apologize for this oversight and the lack of clarity.

Figure 1 & Supplementary Methods (77 samples): The number 116 in the original Figure 1 was a clerical error. The correct number is 77, which is the cohort used for our differential methylation analysis. This number is now consistent with the Supplementary Methods. This cohort is composed of 13 Normal, 15 NAA, 12 AA, and 37 CRC samples. The figure has been revised accordingly.

Abstract (161 samples): The total of 161 plasma samples mentioned in the abstract is the sum of two distinct sample sets used for different stages of our analysis: The 77 samples (13 Normal, 15 NAA, 12 AA, 37 CRC) used for the differential analysis. An additional 84 samples (33 Normal, 51 CRC) which served as the training set for the LASSO regression model. We have now clarified these distinctions in the text and ensured consistency across the abstract, figures, and methods sections.

(b) Methodological inconsistencies

The Supplementary Material reports 477 hypermethylated sites from TCGA data analysis (Δβ>0.20, FDR<0.05), but Figure 1 indicates 499 sites.

The manuscript states that analyzing TCGA data across six cancer types identified 499 CRC-specific methylation sites, yet Figure 1 shows 477. Please also explain the rationale for selecting these specific cancer types from TCGA.

We sincerely thank the reviewer for their sharp observation and for highlighting these inconsistencies. We apologize for this clerical error, which occurred when labeling the figure. The numbers 477 and 499 in Figure 1 were inadvertently swapped and the text in Supplementary Material is correct. We have now corrected this error throughout the manuscript to ensure clarity and consistency. We deeply regret the confusion this has caused.

Regarding the rationale for selecting the cancer types:

The selection of colorectal, esophageal, gastric, lung, liver, and breast cancers was based on the following strategic criteria to ensure the stringent identification of CRC-specific markers. Firstly, esophageal, gastric, liver, and colorectal cancers all originate from the gastrointestinal tract and share developmental and functional similarities. Comparing CRC against these closely related cancers allowed us to filter out general GI-tract-related methylation patterns and isolate those that are truly unique to colorectal tissue. Secondly, we included lung and breast cancer as they are two of the most common non-GI malignancies worldwide with distinct tissue origins. This helps ensure our identified markers are not just pan-cancer methylation events but are specific to CRC, even when compared against highly prevalent cancers from different lineages. Finally, these six cancer types have some of the largest and most complete datasets available in the TCGA database, including high-quality methylation data. This provided a robust statistical foundation for a reliable cross-cancer comparison. We hope this explanation clarifies our methodology. Thank you again for your valuable feedback.

"404 CRC-specific DMRs" mentioned in the main text while "404 MCBs" in Figure 1, the authors need to clarify if these terms are interchangeable or how MCBs are defined.

We sincerely thank the reviewer for pointing out this important inconsistency in terminology. We apologize for the confusion this has caused and for the error in Figure 1. The two terms are closely related in our study. The final 404 markers are technically DMRs that were identified through an analysis of MCBs. To avoid confusion, we have decided to unify the terminology. The manuscript has now been revised to consistently use "DMRs", which is the most accurate final descriptor. The label in Figure 1 has been corrected accordingly.

(2) Methodological documentation

The Results section requires a more detailed description of marker identification procedures and justification of methodological choices.

Figure 3 panels need reordering for sequential citation.

We thank the reviewer for this valuable suggestion. We agree that the original Results section lacked sufficient detail regarding the marker identification procedures and the justification for our methodological choices. To address this, we have substantially rewritten the "Methylation markers selection" subsection. This revised section provides a clear, step-by-step narrative of our marker discovery. The revised text now integrates the specific methodological details and statistical criteria. For instance, we now explicitly describe the three-pronged approach for the initial TCGA data mining and the specific criteria (Δβ, FDR, log2FC) for each, and the analysis methodology such as Wilcoxon test and LASSO regression analysis. We believe this detailed narrative now provides the necessary description and justification for our methodological choices directly within the results, significantly improving the clarity and logical flow of our manuscript. This revision can be found on (Page 9-11, Lines 180-195, 202-213). We hope these changes fully address the reviewer's concerns.

We thank the reviewer for pointing out the citation order of the panels in Figure 3. This was a helpful suggestion for improving the clarity of our manuscript. We have now reordered the panels in Figure 3 to ensure they are cited sequentially within the text. These adjustments have been made in the "Development and validation of the CRC diagnosis model" subsection of the Results (Page 11, lines 224-230). We appreciate the reviewer's attention to detail.

(3) Quality control and data transparency

No quality control metrics are presented for the in-house sequencing data (e.g., sequencing quality, alignment rate, BS conversion rate, coverage, PCA plots for each cohort).

The analysis code should be publicly available through GitHub or Zenodo.

At a minimum, processed data should be made publicly accessible to ensure reproducibility.

We sincerely thank the reviewer for their valuable and constructive feedback regarding quality control and data transparency. We fully agree that these elements are crucial for ensuring the robustness and reproducibility of our research. As the reviewer suggested, we have made all processed data and the key quality control metrics for each sample including sequencing quality scores, bisulfite (BS) conversion rates, and sequencing coverage publicly available to ensure the reproducibility of our findings. The analysis was performed using standard algorithms as detailed in the Methods section. While we are unable to host the code in a public repository at this time, all analysis scripts are available from the corresponding author upon reasonable request. The data has been deposited in the National Genomics Data Center (NGDC) and is accessible under the accession number OMIX009128. This information is now clearly stated in the "Data and Code Availability" section of the manuscript. We thank the reviewer again for pushing us to improve our manuscript in this critical aspect.

Reviewer #3 (Public review):

Summary:

This article provides a model for early diagnosis and prognostic prediction of Colorectal Cancer and demonstrates its accuracy and usability. However, there are still some minor issues that need to be revised and paid attention to.

Strengths:

A large amount of external datasets were used for verification, thus demonstrating robustness and accuracy. Meanwhile, various influencing factors of multiple samples were taken into account, providing usability.

Weaknesses:

There are notable language issues that hinder readability, as well as a lack of some key conclusions provided.

We are very grateful to the reviewer for their positive assessment of our study and for the constructive feedback provided. We are particularly encouraged that the reviewer recognized the strengths of our work, especially the robustness demonstrated through extensive external validation and the practical usability of our model. Regarding the weaknesses, we have taken the comments very seriously and have thoroughly revised the manuscript. We sincerely apologize for the language issues that hindered readability in our initial submission. To address this, the entire manuscript has undergone a comprehensive round of professional language polishing and editing. We have carefully reviewed and revised the text to improve clarity, flow, and grammatical accuracy. Besides, we agree that the conclusions could be stated more explicitly. To rectify this, we have substantially revised the final paragraph of the Discussion and the Conclusion section (Page 14-18, lines 279-305, 319-334, 346-348, 358-360, 367-379). We now more clearly summarize the main findings of our study, emphasize the clinical significance and potential applications of our model, and provide clear take-home messages. We thank you again for your time and insightful comments, which have been invaluable in improving the quality of our paper. We hope the revised manuscript now meets the standards for publication.

Reviewer #1 (Recommendations for the authors):

Detail comments are outlined below:

(1) In this study, the authors have highlighted methylated cfDNA as a noninvasive approach for CRC early diagnosis. However, the small size of cohorts for plasma screening, particularly the sample number of NAA and AA , may cause bias in the selection of DMRs. This bias may lead to inappropriate DMRs for early diagnosis. Furthermore, the similar issues for the training set with a high percentage of late-stage CRC, no AA or NAA samples were included. This absence may be the key factor in screening changed methylated cfDNA that can predict the early stages of CRC.

We are very grateful to the reviewer for this insightful methodological critique. We agree that cohort composition and sample size are critical factors in the development of robust biomarkers, and we appreciate the opportunity to clarify our study design and the interpretation of our results.

We agree with the reviewer that the number of precancerous lesion samples (NAA and AA) in our initial plasma screening cohort was limited. This is a valid point. However, it is important to contextualize the role of this step within our overall multi-stage marker selection funnel. The markers evaluated in this plasma cohort were not discovered from this small sample set alone. They were the result of a rigorous pre-selection process based on large-scale public TCGA data and our own tissue-level sequencing. This robust, tissue-based validation ensured that only the most promising CRC-specific markers were advanced for plasma testing. Therefore, while the plasma cohort was modest in size, its purpose was to confirm the circulatory detectability of markers already known to have a strong tissue-of-origin signal, thereby mitigating the potential bias from a smaller discovery set.

Our primary aim was to first build a model that could robustly and accurately identify a definitive cancer-specific methylation signal. By training the model on clear-cut invasive cancer cases versus healthy controls, we could isolate the most powerful and specific markers for established malignancy. Our working hypothesis was that these strong cancer-specific methylation patterns are initiated during the precursor stages and would therefore be detectable, albeit at lower levels, in precancerous lesions. Unfortunately, the panel could only identify a limited proportion of precancerous lesions (48.4% in the NAA group and 52.2% in the AA group). We fully agree with the reviewer's sentiment that including a larger and more balanced set of precancerous lesions in future training cohorts could potentially optimize a model specifically for adenoma detection. We have now explicitly added this point to our Discussion section, highlighting it as an important direction for future research (Page 18, lines 367-373).

(2) The sensitivity of 27 DMRs in the external validation set (for NAA, AA and CRC 0-Ⅱare 48.4%. 52.2% and 66.7%, respectively) were much lower compared with previously published studies, like ColonES assay (DOI: 10.1016/j.eclinm.2022.101717) and ColonSecure test (DOI: 10.1186/s12943-023-01866-z). The 27 DMRs from the layered screening process did not show superior performance in a small population of an external validation cohort. Therefore, it is unlikely that this DMR pattern will be applicable to the general population in the future.

We sincerely thank the reviewer for their insightful comments and for providing a thorough comparison with the highly relevant ColonES and ColonSecure assays. This has given us an important opportunity to clarify the unique contributions and specific clinical applications of our 27-DMR panel.

We acknowledge the reviewer's point that the sensitivities of our panel for precancerous lesions (NAA: 48.4%, AA: 52.2%), while substantial, are numerically lower than those reported by the excellent ColonES assay (AA: 79.0%). However, it is important to clarify that while the ColonES and ColonSecure tests are outstanding benchmarks designed primarily for early detection and screening, the primary objective and contribution of our study were slightly different. Our model demonstrated an exceptional ability to predict distant metastasis with an AUC of 0.955 and a strong capacity for predicting overall prognosis with an AUC of 0.867. Our goal was to develop a multi-functional, biologically-rooted biomarker panel that not only contributes to early detection but, more importantly, provides crucial information for post-diagnosis patient management, including staging, risk stratification, and prognostication, from a single preoperative sample. We believe this ability to preoperatively identify high-risk patients who may require more aggressive treatment or intensive surveillance is the key contribution of our work. It provides a distinct clinical utility that complements, rather than directly competes with, pure screening assays.

We agree with the reviewer that our external validation was performed on a limited cohort, and we have acknowledged this as a limitation in our Discussion section. However, the purpose of this validation was to provide a proof-of-concept for the panel's performance across its multiple functions. The promising and exceptionally high-performing results in the prognostic domain strongly warrant further validation in larger, prospective, multi-center cohorts.

(3) The 27 DMRs pattern worked well in predicting CRC distant metastasis, and the methylation score remarkably increased in stage III-IV. In contrast, the increase of AA and 0-II groups was very mild in the validation cohort. This observation raises concerns regarding the study design, particularly in the context of the layered screening process and sample assigning.

We sincerely thank the reviewer for this insightful and critical comment. We agree with the reviewer's observation that the methylation score increased more remarkably in late-stage (III-IV) CRC compared to the milder increase in adenoma (AA) and early-stage (0-II) CRC in the validation cohort. However, the observed pattern is biologically plausible and consistent with the nature of colorectal cancer progression. Carcinogenesis is a multi-step process involving the gradual accumulation of genetic and epigenetic alterations. The methylation changes we identified are likely associated with tumor progression and metastasis. Therefore, it is expected that advanced, metastatic cancers (Stage III-IV), which have undergone significant biological changes, would exhibit a much stronger and more robust methylation signal compared to pre-cancerous lesions (adenomas) or early-stage, non-metastatic cancers (Stage 0-II). The "mild" increase in early stages reflects the initial, more subtle epigenetic alterations, while the "remarkable" increase in late stages reflects the extensive changes required for invasion and metastasis. We believe this graduated increase actually strengthens the validity of our methylation signature, as it mirrors the underlying biological progression of the disease. We hope this response and the corresponding revisions address the reviewer's comments.

(4) The authors did not provide the 27 DMRs prediction efficacy comparison with other noninvasive CRC assays, like a CEA and a FIT test.

Thank you for this valuable suggestion. We agree that comparing our model with established non-invasive assays is crucial for demonstrating its clinical potential. Following your advice, we have now included a direct comparison of the diagnostic performance between our model and the traditional tumor marker, carcinoembryonic antigen (CEA), using the external validation cohort. The results show that our model has a significantly higher sensitivity for detecting early-stage colorectal cancer and adenomas compared to CEA. This detailed comparison has been added as Table s7 in the supplementary materials, and the corresponding description has been incorporated into the Results section of our manuscript (Page 12, lines 234-236). Regarding the Fecal Immunochemical Test (FIT), we unfortunately could not perform a direct statistical comparison because very few individuals in our cohort had undergone FIT. A comparison based on such a small sample size would lack statistical power and might not yield meaningful conclusions. We have acknowledged this as a limitation of our study in the Discussion section.We believe these additions and clarifications have substantially strengthened our manuscript. Thank you again for your constructive feedback.

(5) The authors did not explicitly describe how they assigned the plasma samples to the distinct sets, nor did they specify the criteria for the plasma screen set, training set, and validation set. The detailed information for the patient grouping should be listed.

Responce: Thank you for this essential feedback. We agree that a transparent and detailed description of the sample allocation process is crucial for the manuscript. We apologize for the previous lack of clarity and have now revised the Methods section to address this. Our patient cohorts were assigned to the screening, training, and validation sets based on a chronological splitting strategy. Specifically, samples were allocated based on the date of collection in a consecutive manner. This approach was chosen to minimize selection bias and to provide a more realistic, forward-looking assessment of the model's performance, simulating a prospective validation scenario. The screening set comprised 89 tissue samples and 77 plasma samples collected between June to December 2020. The primary purpose of this set was for the initial discovery and screening of potential methylation markers. The training set and validation set included 165 plasma samples collected from December 2020 to July 2022. The external validation cohort comprised 166 plasma samples collected from from July 2022 to December 2022. The subsection titled "Study design and samples" within the Methods section of the revised manuscript, which now contains all of this detailed information (Page 6, lines 116-133). We believe this detailed explanation now makes our study design clear and transparent. Thank you again for helping us improve our manuscript.

Reviewer #2 (Recommendations for the authors):

The manuscript requires significant language editing to improve clarity and readability. We recommend that the authors seek professional editing services for revision.

Thank you for your constructive comments on the language of our manuscript. We apologize for any lack of clarity in the previous version. To address this, we have performed a thorough revision of the manuscript. The text has been carefully reviewed and edited by a native English-speaking colleague who is an expert in our research field. We have focused on correcting all grammatical errors, improving sentence structure, and refining the phrasing throughout the document to enhance readability. We are confident that these extensive revisions have significantly improved the clarity of the manuscript. We hope you will find the current version much easier to read and understand.

Reviewer #3 (Recommendations for the authors):

(1) However, I think the abstract part of the article is too detailed and should be more concise and shortened. It is not necessary to show detailed values but to summarize the results.

Thank you for this valuable suggestion. We agree that the previous version of the abstract was overly detailed and that a more concise summary would be more effective for the reader. Following your advice, we have substantially revised the abstract. We have removed the specific numerical values (such as detailed statistics) and have instead focused on summarizing the key findings and their broader implications (Page 3, lines 54-60, 64-66, 70-72). The revised abstract is now shorter and provides a clearer, high-level overview of our study's background, methods, main results, and conclusions. We believe these changes have significantly improved its readability and impact. We hope you will find the current version more appropriate.

(2) Figure 4, the color in the legend and plot are not the same, and should be revised.

Thank you for your careful attention to detail and for pointing out the color inconsistency in Figure 4. We apologize for this oversight. We have now corrected the figure as you suggested, ensuring that the colors in the legend perfectly match those in the plot. The revised Figure 4 has been updated in the manuscript. We appreciate your help in improving the quality of our figures.

(3) Please pay attention to the article format, such as the consistency of fonts and punctuation marks. (For example, Lines 75 and Line 230).

Thank you for your meticulous review and for pointing out the inconsistencies in our manuscript's formatting. We sincerely apologize for these oversights and any inconvenience they may have caused. Following your feedback, we have carefully corrected the specific issues you highlighted. Furthermore, we have conducted a thorough proofread of the entire manuscript to ensure consistency in all fonts, punctuation marks, and overall adherence to the journal's formatting guidelines. We appreciate your help in improving the presentation and professionalism of our paper.

Version 2

Abstract

Cell-free DNA (cfDNA) methylation is a valuable biomarker in various cancers including colorectal cancer (CRC), but marker for both early diagnosis and prognostic prediction remain a critical unmet need. Here, we report the development of a 27-DMR (differentially methylated regions) plasma panel with dual diagnostic and prognostic functions. We first identified CRC-specific methylation features from 119 tissue samples and 161 plasma samples. Machine learning algorithms were then applied to develop diagnosis and prognosis models using the plasma samples in training cohort. In the tissue external validation cohort (GSE48684), the cfDNA methylation diagnosis model conducted with the panel, achieved an area under the curve (AUC) of 0.983, and for the plasma cfDNA model in the external validation cohort, the sensitivities for NAA, AA and CRC 0 - II are 48.4%, 52.2% and 66.7% respectively, with a specificity of 88%. Beyond its diagnostic capabilities, the methylation panel was also a powerful predictor of distant metastasis (AUC = 0.955) and patient prognosis (AUC = 0.867). Using normal samples as control, the changes in methylation score in both tissue and plasma were consistent across different lesions, although the degree of alterations varied with severity. The methylation scores vary between paired tissue and blood samples, suggesting distinct mechanisms of migration from tumor tissue to blood for the 27 DMRs. Together, our cfDNA methylation models based on 27 DMRs can identify different stages of CRC and predict metastasis and prognosis, ultimately enabling early intervention and risk stratification for CRC patients.

Introduction

Colorectal cancer (CRC) is the third most prevalent malignancy and the fourth leading cause of cancer-related mortality worldwide, accounting for 9.4% of all cancer deaths (1). When patients were diagnosed with CRC, over half of their family (62.9%) faced financial burdens (2, 3). The survival rates of individuals diagnosed with advanced CRC (Stage III or IV) witness a significant decrement (4, 5), underscoring early detection and removal of precancerous lesions remain the most effective strategy to prevent CRC-associated mortality (6). Although colonoscopy is the gold standard for CRC detection, its invasiveness, suboptimal patient compliance and risk of complications including intestinal perforation limit its widespread effectiveness (7). Other non-invasive tests, such as guaiac-based fecal occult blood test (gFOBT), fecal immunochemical test (FIT), carcino-embryonic antigen (CEA), and multi-target stool DNA (mt-sDNA) testing, are constrained by insufficient sensitivity, particularly for early-stage lesions (8, 9).

Recent research demonstrated a profound correlation between CRC and the development of genetic and epigenetic alterations. During early stage of tumorigenesis, epigenetic modifications surpass the frequency of gene mutations, indicating their potential as diagnostic biomarkers in screening for colon polyps and cancer (10). Circulating tumor DNA (ctDNA), a component of cell-free DNA (cfDNA) released from apoptotic and necrotic tumor cells, carries cancer-specific epigenetic alterations (11). Consequently, cfDNA methylation emerges as a promising biomarker for cancer screening due to its early onset and high signal density(12).

While the utility of cfDNA methylation in early diagnosis, surveillance, molecular subtyping and prognostic prediction of CRC has been proven, current approaches have limitations (13). For instance, a diagnostic panel has shown excellent performance but was developed using inconsistent sample types (e.g., CRC tissue vs. normal blood leukocyte methylation data), introducing potential bias (14). Another study addressed the potential of cfDNA methylation in patient risk stratification, but required blood sampling every three months, which can hamper patient compliance and clinical feasibility (15).

In this study, we aimed to develop and validate a plasma-based biomarker panel, initially identified from CRC tissues, for the comprehensive diagnosis, metastatic assessment, and prognostic prediction of CRC. We performed a meticulous, multi-stage screening process starting with CRC and normal tissues, followed by refinement using plasma samples. Crucially, by utilizing a substantial cohort of paired tissue and plasma samples from the same patients, we minimized potential confounding biases. The resulting 27 differential methylated regions (DMRs) panel was designed to, from a single blood test, ascertain the presence of CRC, distinguish lesion stages, identify distant metastasis, and stratify patients by risk. Such a multi-functional tool holds significant promise for guiding the clinical management of CRC, from early detection to personalized treatment strategies.

Materials and methods

Study design and samples

This study was designed to first identify CRC-specific DNA methylation markers and then to develop and validate models for CRC diagnosis, metastasis, and prognosis. The marker discovery phase utilized data from TCGA public database and clinically collected tissue and plasma samples. The 450k chip methylation data encompassing colorectal, esophageal, gastric, lung, liver, and breast cancers, and CRC transcriptome sequencing data were downloaded from The Cancer Genome Atlas (TCGA) database. The methylation data of tissue screening cohort was obtained from fifteen pairs of tissue samples sequenced with the Roche Nimble Gen Seq Cap Epi method from the Seventh Medical Center of PLA General Hospital. The 89 tissue samples and 77 plasma samples used for marker screening were collected between June to December 2020 at the Seventh Medical Center of PLA General Hospital and Dongying People’s Hospital. Participants for model establishment and validation were enrolled from December 2020 to July 2022 in the Seventh Medical Center of PLA General Hospital, Dongying People’s Hospital. The paeticipants for external validation cohort were enrolled from July 2022 to December 2022 at the First Hospital of Longyan, Fujian Medical University. An additional public tissue validation cohort (GSE48684) was obtained from the Gene Expression Omnibus (GEO). Detailed patient inclusion and exclusion criteria, as well as methods for DNA extraction, library construction, and targeted bisulfite sequencing, are provided in the Supplementary Materials.

CRC-specific methylation markers selection

An initial marker pool was generated by integrating TCGA methylation data with our 15 pairs of tissue methylation sequencing results. This informed the design of a targeted panel, which was then screened in the 89 samples (13 normal tissues and 38 paired CRC tissues), identifying 404 DMRs. Then, the candidates were further evaluated in 77 plasma samples from 13 Normal, 15 non-advanced adenoma (NAA), 12 advanced adenoma (AA) and 37 CRC participants. Subsequently, the Least Absolute Shrinkage and Selection Operator (LASSO) regression analysis was used to filter markers, and 27 regions that appeared in more than 50 iterations were chosen as plasma diagnostic markers. The correlation between the methylation levels of the 27 DMRs and the expression of their associated genes was assessed using Spearman’s rank correlation.

Construction of the CRC diagnosis model

A diagnostic model was developed using a logistic regression algorithm trained on methylation data from the training cohort. The model was built using five-fold cross-validation, optimizing for the minimum binary deviance. Its performance was evaluated by Receiver Operating Characteristic (ROC) curve analysis and the optimal cutoff value for the resulting methylation score was determined using the maximum Youden index. Using the methylation scores derived from the diagnosis model, the CRC staging (AA, CRC 0-II, CRC III-IV) was predicted.

Construction of the CRC metastasis and prognosis models

A metastasis prediction model was constructed using the methylation score from the diagnostic model combined with patients’ clinical metastasis status. Similarly, a prognostic model was developed by integrating the methylation score with patient survival data. The performance of both models was evaluated using ROC analysis, with optimal cutoffs determined by the maximum Youden index.

Statistical analysis

All statistical analyses were performed using SPSS 26.0 or R 4.1.3. Two-sided tests were used for p-values, with differences deemed statistically significant at p < 0.05. Model performance was assessed by ROC analysis using the “roc” function from the R package pROC, generating the area under the ROC curve (AUC) with 95% confidence interval (CI). Survival analyses were conducted using Kaplan-Meier (KM) curves with the log-rank test and hazard ratios were calculated using Cox proportional-hazards models.

Results

Patient characteristics

To delineate DNA methylation biomarkers specific to CRC, 119 tissues and 77 plasma samples (originating from the same patients with tissues) were collected for methylation sequencing analysis, narrowing down to 142 DMRs. Then the 27 DMRs were selected with the sequencing data of 84 plasma samples. Following this, CRC diagnosis, metastasis, and prognosis models were constructed using plasma samples from 84 individuals (51 CRC, 33 Normal). Validation was then carried out in an independent cohort (30 CRC, 37 AA, 14 Normal). The CRC diagnosis model was further tested in an external validation cohort (18 CRC 0-II, 91 NAA, 23 AA, 34 Normal). Detailed information about the study design is illustrated in Figure 1, and comprehensive patient characteristics are summarized in Table s2-s4.

Flow diagram of the study.

Methylation markers selection

We employed a multi-stage funneling strategy to identify a robust panel of CRC-specific methylation markers, starting from a broad analysis of public data and progressively refining the list through validation in our own tissue and plasma samples. Analyzing the 450k chip methylation data from TCGA, we identified the top 500 differentially methylated CpG sites (DMCs) between CRC and control tissues (Δβ > 0.50, FDR < 0.05), 477 hypermethylated sites (Δβ > 0.20, FDR < 0.05) located in the TSS or First Exon regions of significantly downregulated genes (log2FC < −1, FDR < 0.05), and 499 sites that were hypermethylated in CRC (Δβ > 0.20) but not in five other major cancers (esophageal, gastric, lung, liver, and breast cancer; Δβ ≤ 0.20). The union of these three approaches, supplemented with known CRC-related methylation sites from existing literature, yielded an initial list of 1,438 candidate DMCs (Supplementary Figure 1A). Unsupervised clustering of these sites revealed distinct methylation patterns between normal and tumor tissues in the TCGA cohort (Figure s1B). To validate these findings in our patient population, we performed methylation sequencing (SeqCap Epi) on 15 paired tissue samples (5 advanced adenoma pairs and 10 CRC pairs). From this analysis, we selected the top 1,400 DMCs based on their statistical significance (Δβ > 0.20, FDR < 0.05) (Figure s2A-C). The distribution of the DMCs was predominantly observed in introns and promoters, signifying substantial implications for gene expression and cellular functions (Figure s2D). Merging the 1438 DMCs with the 1400 DMCs for probe design, 404 DMRs were discerned across 89 tissues (13 normal tissues and 38 paired CRC tissues) (Figure s3A). The different methylation patterns between Normal and Tumor were also quite evident (Figure s3B-C). These DMRs correspond to genes that play pivotal roles in biological processes such as cell-cell adhesion, digestive system development, and cAMP and cGMP pathways (Figure s3D). To identify markers detectable non-invasively, we assessed the 404 DMRs in 77 plasma samples (13 Normal, 15 NAA, 12 AA, 37 CRC) and narrowed the list to 142 DMRs that showed significant differential methylation in plasma (Wilcoxon test; Δβ > 0.01, FDR < 0.05) (Figure s3E-F). Finally, to select the most robust and minimal signature for diagnostic modeling, we applied a Least Absolute Shrinkage and Selection Operator (LASSO) regression analysis to the training cohort methylation data. The analysis involved 500 random iterations, and we selected markers that appeared in more than 50 iterations and yielded a final panel of 27 top-performing DMRs for plasma-based CRC detection. Detailed information on these 27 DMRs is provided in Table s5. Correlation analysis using TCGA data confirmed that the methylation levels of the majority of these 27 DMRs were significantly correlated with the expression of their associated genes, underscoring their biological relevance (Figure s4).

Development and validation of the CRC diagnosis model

Using methylation data from 27 DMRs in tissues, we applied machine learning to construct a diagnosis model for CRC and the AUC reached a high value of 0.994 (Figure 2A). In the external tissue independent validation cohort (GSE48684), the AUC for CRC is 0.983 and for AA is 0.966. (Figure 2B-C). The sensitivities for AA and CRC were 95% and 94%, respectively (Figure 2D-E). Besides, the plasma-based diagnosis model for CRC was constructed with training cohort and validated in an external independent validation cohort, the 27 DMRs exhibited distinctive methylation patterns between CRC and normal individuals (Figure 3A-B). The methylation scores of the model were significantly elevated in the CRC group compared to the normal and AA groups (P< 0.001), and they increased with advance tumor staging (Figure 3C). In the training cohort, the ROC curve analysis showed an AUC value of 0.928 [95% confidence interval (CI): 0.874–0.981] (Figure 3D). ROC curve analysis in the validation cohort showed AUC values for AA, stages 0-II, III-IV, and all CRC stages as 0.714 (0.550–0.878), 0.890 (0.749–1.000), 0.967 (0.899–1.000), and 0.929 (0.827–1.000).

Development and validation of tissue diagnosis model based on the 27 CRC-specific DMRs.

(A) The use of ROC curve analysis to assess the performance of the tissue diagnostic model in differentiating CRC patients from normal individuals of the training cohort (51 tissues we collected). (B-C) The use of ROC curve analysis to assess the performance of the tissue diagnostic model in differentiating CRC patients from normal individuals(B) and distinguish AA from Normal subjects (C) in tissue validation cohort 1(GSE 48684 dataset). (D-E) The sensitivity and specificity of the diagnosis model in the tissue training(D) and validation cohort 1(E).

Development and validation of a plasma diagnosis model based on the 27 CRC-specific DMRs.

(A-B) Heatmap illustrating the DMRs between CRC and advanced adenoma and healthy controls in the training (A) and validation cohort (B). (C) Methylation scores of Normal、AA and CRC individuals generated from the diagnosis model in the training and validation cohort (The dashed line represents the cutoff value). (D)The use of ROC curve analysis to assess the performance of the diagnosis model in differentiating CRC patients from normal individuals of the training cohort. (E) The use of ROC curve analysis to evaluate the performance of the diagnosis model in differentiating CRC patients from normal individuals of the validation cohort. (F-H) In the validation cohort, the performance of the diagnosis model in differentiating early-stage CRC (0-II) from normal individuals (F); in differentiating advanced-stage CRC (III-IV) from normal individuals (G); in differentiating AA from normal individuals (H). (I) The sensitivity and specificity of the diagnosis model stratified by stages in the training and validation cohort. Abbreviations: DMRs, differential methylated regions; ROC, receiver operating characteristic; AA, advanced adenoma; CRC, colorectal cancer.

(Figure 3E-H). The sensitivities of the CRC diagnosis model in the training cohort for stages 0-II, III-IV, and all CRC stages were 80%, 92%, and 86%, respectively, with a specificity of 85% (Figure 3I). The CRC diagnosis model demonstrated a specificity of 93% in the validation cohort, with sensitivities for AA, stages 0-II, III-IV, and all CRC stages of 43%, 67%, 100%, and 83%, respectively (Figure 3I). We further tested the early diagnostic performance of the model in an external validation cohort, and the results showed that the model achieved a sensitivity of 52% for AA and 48% for NAA (Figure 4A-E). Compared to the traditional tumor marker CEA, the model demonstrated higher sensitivity in detecting early-stage colorectal cancer and adenomas (Table s7). The plasma diagnosis model also serves to discern the specific staging of CRC patients. ROC curve analysis on the validation cohort revealed an AUC of 0.813 (0.656–0.970) for discriminating between CRC 0-II and CRC III-IV, and an AUC of 0.799 (0.693–0.905) for distinguishing between AA and CRC (Figure s5A-C). Kaplan-Meier survival curves revealed a significant reduction in overall survival (OS) for CRC III-IV patients compared to CRC 0-II patients identified by this diagnosis model (Figure s5D-E).

External validation of plasma diagnosis model.

(A-C) In the external validation cohort, the performance of the diagnosis model in differentiating early-stage CRC (0-II) from normal individuals (A); in differentiating AA from normal individuals (B); in differentiating NAA from normal individuals (C). (D) Methylation scores of Normal、NAA、AA and CRC (0-II) individuals generated from the diagnosis model in the external validation cohort (The dashed line represents the cutoff value). (E) The sensitivity and specificity of the diagnostic model for NAA, AA, and CRC in the external validation cohort.

Development and validation of CRC metastasis and prognosis models

To investigate the clinical relevance of our methylation signature, we first correlated the plasma methylation scores with clinicopathological parameters. The scores were significantly associated with tumor stage and metastasis (P < 0.001, Figure 5A-B) but not with age, sex, or lesion location (P > 0.05, Figure s6A-C). Based on these associations, we developed models to predict metastasis and prognosis using the 27-DMR panel. The metastasis model demonstrated excellent performance, achieving an AUC of 0.969 (95% CI: 0.926-1) in the training cohort and 0.955 (95% CI: 0.878-1) in the validation cohort (Figure 5C). Correspondingly, patients with metastatic disease (M1) had significantly higher methylation scores than those without (M0) and exhibited shorter overall survival (OS) (Figure s6D-F). The prognosis model also showed high accuracy, with AUCs of 0.883 (95% CI: 0.778-0.988) in the training set and 0.867 (95% CI = 0.728-1) in the validation set (Figure 5D). Using an optimal cutoff, the prognosis model effectively stratified patients into high-and low-risk groups. Kaplan-Meier analysis revealed that patients in the high-risk group had significantly worse OS (Figure 5E). Critically, this high-risk group encompassed nearly all patients who developed distant metastases or died during follow-up (Figure 5F-I). To confirm the score’s independent prognostic value, we performed a multivariate Cox regression analysis, which showed that the 27-DMR methylation score was an independent predictor of OS (Table S6). These findings establish that our models can successfully stratify patients, identifying those who may require more aggressive treatment.

Development and validation of a plasma-based metastasis model and a plasma-based prognosis model.

(A-B) The methylation scores generated by the diagnosis model for CRC patients across various M stages (A) and different disease stages (B). (C) ROC curve analysis was performed to assess the performance of 27 DMRs methylation scores generated by the diagnosis model in distinguishing M1-CRC patients from M0-CRC patients in the training (AUC = 0.969) and validation (AUC = 0.955) cohorts. (D) ROC analysis evaluates the performance of 27 DMRs methylation scores generated by the diagnosis model in predicting the prognosis of CRC patients. (E) Kaplan-Meier survival curves comparing OS between the high-risk CRC group and low-risk CRC group of prognosis model in the training cohort and validation cohort. (F, H) Metastasis and prognosis prediction of the training and validation patitients (n = 88). The black dotted line at the cutoff value divides the patients into high-risk and low-risk groups. Yellow circles and gray circles represent patients without distant metastasis and with distant metastasis, respectively (F). Yellow circles and gray circles represent patients survived and deceased, respectively (H). (G, I) The proportion of metastasis and deceased is higher in the high-risk group. The p-value was calculated using a two-sided Fisher’s exact test. Abbreviations: DMRs, differential methylated regions; ROC, receiver operating characteristic; M1-CRC, colorectal cancer with distant metastasis; M0-CRC, colorectal cancer without distant metastasis; AUC, the area under the curve.

Distinct Release Dynamics of Methylation Markers from Tissue to Plasma

To understand the biological basis for the model’s performance characteristics, we compared methylation scores between paired tissue and plasma samples across the spectrum of disease progression. We found that in tissue, the 27-DMR signature was already strongly present at the precancerous adenoma (AA) stage (Figure 6A). However, in plasma, the signal intensity increased in a stepwise manner, becoming most prominent in advanced-stage CRC (Figure 6B). This discrepancy suggests that the release of cfDNA markers into circulation is stage-dependent. Indeed, a subset of DMRs was detectable in plasma even in early-stage disease, and these markers were associated with genes involved in cellular secretion pathways (Figure 6C, s6G). In contrast, other DMRs were only detected in the plasma of patients with advanced CRC. This suggests a dual mechanism of release: limited, active secretion of certain markers from early lesions, followed by massive, passive release from widespread cell death in larger, necrotic tumors. This dynamic explains why the diagnostic sensitivity of the plasma test is highest for advanced-stage disease.

Methylation scores of paired tissue and plasma samples.

(A) The methylation scores of tissues from Normal, AA, CRC 0-II, CRC III-IV individuals. (B) The methylation scores of blood samples from Normal, AA, CRC 0-II, CRC III-IV individuals. (C) The mechanism diagram interpreting the methylation scores changes of tissue and plasma across different populations.

Discussion

Liquid biopsy of cfDNA has become an ideal clinical detection method because it is minimally invasive and easily sampled. However, a key challenge in liquid biopsy is confirming that cfDNA markers truly originate from tumor tissue. We addressed this by implementing a rigorous, multi-stage screening strategy that began with TCGA data, was validated in our own CRC tissue samples, and was finally optimized in plasma. This ensures that our 27-DMR signature is not only statistically powerful but also biologically rooted in CRC pathobiology. In this study, we implemented a rigorous, multi-stage screening strategy that began with integrating TCGA tissue data with self-collected tissue methylation data to identify CRC-related methylation sites, was validated in our own CRC tissue samples, and was finally optimized in plasma. This ensures that our 27-DMR signature is not only statistically powerful but also biologically rooted in CRC pathobiology. Our resulting diagnostic model, built upon these tissue-of-origin-confirmed markers, demonstrates significant potential for early diagnosis, staging, metastasis and prognostic assessment.

A critical benchmark for any new CRC screening tool is its performance against existing methods, especially in detecting precancerous lesions. For instance, the FDA-approved circulating methylated SEPT9 DNA (mSEPT9) offers a specificity of 91.5% but provides limited sensitivities of 11.2%, 35.0%, 63.0%, 46.0%, and 77.4% for AA and CRC stages I-IV, respectively (16). While more recent cfDNA methylation models have improved overall CRC detection, their ability to detect precancerous lesions can still be modest; one such study reported a sensitivity of 33.3% for adenomas (17). In this context, our compact 27-marker plasma model demonstrates a highly competitive performance, achieving an overall sensitivity of 83.3% and a specificity of 92.9% for CRC. Most notably, our model shows a substantial improvement in detecting high-risk precancerous lesions, with sensitivities for advanced adenomas (AA) and non-advanced adenomas (NAA) reaching 52.2% and 48.4%, respectively. This represents a nearly five-fold increase in sensitivity for advanced adenomas compared to the mSEPT9 test (16, 17). This enhanced capability positions our model as a promising new tool for CRC screening in high-risk populations, enabling earlier intervention before the onset of malignancy.

In plasma, we observed the methylation scores of 27 DMRs gradually increase with CRC progression. It is possibly due to less vascular infiltration in early CRC (18, 19). On the other hand, the ctDNA detected in the blood of early-stage CRC is largely derived from tumor cells actively secreting into the bloodstream, making it challenging to be precisely captured with current detection technologies due to the limited quantity. Conversely, ctDNA in the blood of late-stage CRC primarily emanated from apoptotic and necrotic tumor cells, resulting in a larger quantity that facilitates easier detection. However, in reality, the methylation scores of the 27 DMRs have indeed exhibited noticeable changes in AA and early cancerous tissues. Additionally, in the tissue diagnosis model using the 27 DMRs, the sensitivity was 94% and the specificity was 100%. Therefore, we have grounds to believe that the 27 DMRs detected in plasma originate from CRC tumor tissues. With the implementation of more sensitive detection methods, the performance of these 27 DMRs in early CRC plasma diagnosis is likely to be further enhanced.

Beyond early detection, cfDNA analysis is a valuable tool for risk stratification and monitoring for recurrence in CRC (20). However, many current approaches require serial postoperative testing to track disease progression (15, 21). A key advantage of our model is its ability to provide significant prognostic information from a single, preoperative blood sample. Our findings reveal that the 27-DMR methylation score progressively increases with CRC stage advancement, a feature with direct clinical utility (22, 23). Firstly, this allows for effective preoperative staging, as the model can distinguish early-stage (0–II) from late-stage (III–IV) CRC with high accuracy (AUC = 0.813), offering a simple, non-invasive method to help guide initial treatment planning. Secondly, and perhaps more significantly, the preoperative score is a powerful predictor of patient prognosis. It was strongly associated with distant metastasis (AUC = 0.955), enabling the stratification of patients into high-and low-risk groups before surgery even begins. This capability is particularly valuable as it provides crucial prognostic insights that can inform perioperative management—such as identifying patients who may require more aggressive treatment or intensive surveillance—from a single baseline measurement, potentially reducing the reliance on continuous postoperative monitoring.

The treatment strategies for CRC differ significantly across various stages. Molecular stratification approaches for CRC patients are on the rise, and concurrently, the clinical application of biomarkers to determine treatment decisions is gradually gaining traction (24). Studies have demonstrated the precise identification of T1 CRC patients at risk of lymph node metastasis using a specific set of miRNAs, thereby potentially mitigating unnecessary overtreatment (25, 26). Our CRC diagnosis model, with an AUC of 0.813 for distinguishing CRC 0-II from CRC III-IV, could also serve as a potent, straightforward, and cost-effective preoperative screening/detection method, guiding patients to select more appropriate treatment plans.

DNA methylation regulates gene transcription, guiding the progression from normal mucosa to AA and ultimately CRC. This process involves silencing tumor suppressor genes and activating oncogene transcription (27). The clinical performance of our panel is underpinned by the biological functions of its associated genes, many of which are known to be involved in CRC pathogenesis. For instance, high methylation and diminished expression of transmembrane protein 240 (TMEM240) regulate CRC cell proliferation, predicting poor prognosis (28). The transcription factor homeobox A3 (HOXA3) activates aerobic glycolysis, promoting tumor growth (29). KIFC3 controls mitotic spindle assembly initiation (30). Moreover, several genes are implicated in tumor metastasis, such as NOVA alternative splicing regulator 1 (NOVA1), which promotes CRC migration by activating the Notch pathway (31). Protein tyrosine phosphatase receptor type T (PTPRT) contributes to early CRC dissemination (32). SPARC-related modular calcium binding 2 (SMOC2) serves as the distinctive signature of cancer stem cells (CSCs) in CRC and promotes epithelial-to-mesenchymal transition (EMT) (33, 34). Increased methylation and expression of pancreatic and duodenal homeobox 1 (PDX1) facilitate CRC invasion and migration(35). The correlation we observed between the methylation levels of our 27 DMRs and the transcription levels of their respective genes in TCGA datasets further supports their functional relevance. However, the mechanisms by which alterations in specific DNA methylation impact gene expression are intricate. Certain transcription factors selectively recognize sequences with methylated CpG (mCpG) and influence the expression of multiple genes(36). Further exploration of the potential functional mechanisms of these DNA methylation markers may deepen our understanding of the molecular processes underlying CRC development, offering promising therapeutic targets. Despite the promising results, we acknowledge several limitations in this study. Firstly, although obtained from multiple institutions, the overall sample size is relatively small. Therefore, the cfDNA methylation model should be further validated in larger, prospective clinical trials including a larger and more balanced set of precancerous lesions. Secondly, a prospective study is necessary to directly compare or combine the cfDNA methylation model with clinically established protein markers such as CEA and CA19-9, and Fecal Immunochemical Test (FIT) to determine its ultimate clinical utility.

In conclusion, our study introduces a novel, tissue-validated 27-DMR cfDNA methylation signature that effectively serves three critical clinical needs: early detection of CRC, including high-risk precancerous lesions; preoperative staging; and prognostic risk stratification for metastasis. As a robust, convenient, and cost-effective detection method, the preoperative application of this biomarker panel has the potential to facilitate more informed clinical decision-making and significantly improve the management of patients with colorectal cancer.